---
title: "HW 2"
author: Malvika Venkataraman
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

# Problem 1

## Read in the Trash Wheel Dataset and Clean
```{r}
trash_wheel_path = "./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx"

trash_wheel_df = read_excel(trash_wheel_path, sheet = "Mr. Trash Wheel", range = "A2:N534") %>%
  janitor::clean_names() %>%
  filter(dumpster != "NA") %>%
  mutate(sports_balls = as.integer(sports_balls))

#view dataset
#View(trash_wheel_df)
```

## Read in the Precipitation Data (2018 and 2019) and Clean

### Read and clean 2018 precipitation data
```{r}
precip_data_2018 = read_excel(trash_wheel_path, sheet = "2018 Precipitation", range = "A2:B14") %>%
  janitor::clean_names() %>%
  filter(total != "NA") %>%
  mutate(year = 2018)

#View(precip_data_2018)
```

### Read and clean 2019 precipitation data
```{r}
precip_data_2019 = read_excel(trash_wheel_path, sheet = "2019 Precipitation", range = "A2:B14") %>%
  janitor::clean_names() %>%
  filter(total != "NA") %>%
  mutate(year = 2019)

#View(precip_data_2019)
```

### Combine 2018 and 2019 datasets
```{r}
precip_data = bind_rows(precip_data_2018, precip_data_2019) %>%
  mutate(month = month.name[month]) %>%
  select(year, month, total)

#View(precip_data)
```

## Interpretation

In the Mr. Trash Wheel dataset, there are `r nrow(trash_wheel_df)` rows and `r ncol(trash_wheel_df)` columns. There are `r max((pull(trash_wheel_df, dumpster)))` observations between 5/16/14 and 1/4/21, for the seven categories of trash represented in the dataset. Key variables include the Weight of the trash (in tons), and the Volume (in cubic yards). The median number of sports balls in a dumpster in 2019 was `r median(pull(trash_wheel_df %>% filter(year == 2019), sports_balls))`.

In the 2018 and 2019 Merged Precipitation dataset, there are `r nrow(precip_data)` rows and `r ncol(precip_data)` columns. There are `r nrow(precip_data)` observations between 1/2018 and 12/2019. Key variables include the month, year and the total precipitation, for each month. For the available data, the total precipitation in 2018 was `r sum(pull(precip_data_2018, total))`

# Problem 2

## Read in the FiveThirtyEight Datasets and Clean

### File Paths
```{r}
pols_month_path = "./data/fivethirtyeight_datasets/pols-month.csv"
unemployment_path = "./data/fivethirtyeight_datasets/unemployment.csv"
snp_path = "./data/fivethirtyeight_datasets/snp.csv"
```

### Pols Month Dataset
```{r}
#read and clean pols-month.csv
party = c("dem","gop")

pols_data =
  read_csv(pols_month_path) %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year","month","day")) %>%
  mutate(month = month.name[as.integer(month)],
         president = party[prez_gop + 1]) %>%
  select(year, month, president, everything(), -day, -prez_dem, -prez_gop)

#view completed dataset
#View(pols_data)
```

### SNP Dataset
```{r}
#read and convert dates to full year to match the pols month and unemployment csv's
snp_data <- read_csv(snp_path)
dates <- snp_data$date
dates <- as.Date(dates, "%m/%d/%y")
dates <- as.Date(ifelse(dates > Sys.Date(), format(dates, "19%y-%m-%d"), format(dates)))
snp_data[,"date"] <- dates 

#clean snp.csv
snp_data =
  snp_data %>%
  janitor::clean_names() %>%
  separate(date, into = c("year","month","day")) %>%
  mutate(month = month.name[as.integer(month)]) %>%
  select(year, month, snp_index = close)

#view completed dataset
#View(snp_data)
```

### Unemployment Dataset
```{r}
#read and clean unemployment csv
unemployment_data =
  read_csv(unemployment_path) %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "percentage") %>%
  mutate(month = month.name[match(pull(.,month),month.abb)],
         Year = as.character(Year)) %>%
  select(year =  Year, month, unemployment_percentage = percentage) #change name to make it more clear

#view completed dataset
#View(unemployment_data)
```

### Merge the Datasets
```{r}
#merge
fivethirtyeight_merged_data =
  left_join(pols_data, snp_data, by = c("year","month")) %>%
  left_join(unemployment_data, by = c("year","month"))

#view completed dataset
#View(fivethirtyeight_merged_data)
```

### Results

The `pols_data` has `r nrow(pols_data)` rows and `r ncol(pols_data)` columns. It indicates, for a specific year and month, whether the president was a republican (`gop`) or a democrat (`dem`). It also indicates the number of republican or democratic governors (`gov_gop`, `gov_dem`), as well as representatives (`rep_gop`, `rep_dem`). The range of years the dataset includes is `r pull(pols_data, year)[1]` to `r rev(pull(pols_data, year))[1]`.

The `snp_data` has `r nrow(snp_data)` rows and `r ncol(snp_data)` columns. It indicates the closing values of the S&P stock index at the beginning of a designated year and month. Key variables include the year, month and S&P index.
The range of years the dataset includes is `r pull(snp_data, year)[1]` to `r rev(pull(snp_data, year))[1]`.

The `unemployment_data` has `r nrow(unemployment_data)` rows and `r ncol(unemployment_data)` columns. It indicates, the unemployment percentage at the beginning of a designated year and month. Key variables include the year, month and unemployment percentage. The range of years the dataset includes is `r pull(unemployment_data, year)[1]` to `r rev(pull(unemployment_data, year))[1]`.

The resulting dataset (`fivethirtyeight_merged_data`) has `r nrow(fivethirtyeight_merged_data)` rows and `r ncol(fivethirtyeight_merged_data)` columns. The years in the dataset range from `r pull(fivethirtyeight_merged_data, year)[1]` to `r rev(pull(fivethirtyeight_merged_data, year))[1]`. Key variables include the year and the month. The resulting dataset provides information on the political party of the president, the number of republican or democratic governors, senators and representatives, the closing values of the S&P stock index, and the unemployment percentage for a specific year and month.

# Problem 3

## Load baby name dataset
```{r}
#load, read, view data
baby_name_path = "./data/Popular_Baby_Names.csv"

baby_name_df = read_csv(baby_name_path)

#View(baby_name_df)
```

## Check for issues
```{r}
unique(baby_name_df[c("Year of Birth")]) #no issues
unique(baby_name_df[c("Gender")]) #no issues
unique(baby_name_df[c("Ethnicity")]) #issues

```

## Tidy the data
```{r}
baby_name_df =
  read_csv(baby_name_path) %>%
  janitor::clean_names() %>%
  distinct() %>%
  mutate(
    gender = str_to_lower(gender),
    ethnicity = str_to_lower(ethnicity),
    childs_first_name = str_to_lower(childs_first_name),
    ethnicity = replace(ethnicity, ethnicity == "asian and paci",
                        "asian and pacific islander"),
    ethnicity = replace(ethnicity, ethnicity == "black non hisp",
                        "black non hispanic"),
    ethnicity = replace(ethnicity, ethnicity == "white non hisp",
                        "white non hispanic")
    )

#view completed dataset
#View(baby_name_df)
```

## Table: Popularity of Name Olivia
```{r}
olivia_table_df =
  filter(baby_name_df, gender == "female", childs_first_name == "olivia") %>%
  select(year_of_birth, ethnicity, rank) %>%
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "rank"
  )

knitr::kable(olivia_table_df, format = "html", 
             caption = 
               "Table 1: Popularity Rank of 'Olivia' as a Female Baby Name Over Time")
```

## Table: Most Popular Name Among Male Children
```{r}
pop_male_name_df =
  filter(baby_name_df, gender == "male", rank == "1") %>%
  select(year_of_birth, ethnicity, childs_first_name) %>%
  pivot_wider(
    names_from = "year_of_birth",
    values_from = "childs_first_name"
  )

knitr::kable(pop_male_name_df, format = "html", 
             caption = 
               "Table 2: All Time Most Popular Name Among Male Children")
```

## Scatterplot
```{r}
name_popularity_plot_df = 
  filter(
    baby_name_df,
    year_of_birth == "2016",
    gender == "male",
    ethnicity == "white non hispanic") %>%
  select(rank, count)

#plot
ggplot(name_popularity_plot_df, aes(x = rank, y = count)) +
  geom_point() +
  labs(
    title = 
      "Plot of Name Popularity for White, Non-Hispanic, Male Children Born in 2016",
    x = "Rank in Popularity of a Name",
    y = "Number of Children With a Name"
  )
```

